---
title: "Multigroup mixture models for performance decline with MGmixIRT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multigroup mixture models for performance decline with MGmixIRT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

## Background

In low-stakes educational assessments some test takers do not sustain
effort to the end of the test: the probability of a correct response
drops for end-of-test items. Ignoring this *performance decline* (PD)
biases item parameters and understates proficiency, and — because groups
can differ in how many students show PD — it can distort group
comparisons, the core product of large-scale assessment programmes
(List, Robitzsch, Lüdtke, Köller, & Nagy, 2017).

**MGmixIRT** estimates the three mixture IRT models of PD compared by
List et al. (2017), each extended to $G \ge 2$ known groups:

* **2PDM** (Bolt, Cohen, & Wollack, 2002): one decline class per group;
  after a pre-specified switching point $i_0$, item intercepts drop.
* **HYBRID** (Yamamoto, 1995): each item position is a potential
  switching point; after it, responses no longer depend on ability.
* **MPDM** (Jin & Wang, 2014): after the switching point ability is
  reduced by a decrement that grows the earlier the onset.

In the multiclass models the decline-class probabilities follow the
parsimonious shape-parameter function of Cao and Stokes (2008), with a
group-specific shape $\omega_g$ and no-decline proportion $\pi_{Ig}$;
class-specific ability means follow a linear function with slope
$\rho_g$. The reference (first) group's no-decline ability distribution
is standardised for identification.

## Quick start

Simulate two groups with different amounts of decline and fit the
generating model:

```{r fit}
library(MGmixIRT)
sim <- sim_mgmixirt("hybrid", n = c(400, 400), I = 8,
                    pi_nd = c(0.50, 0.70), omega = c(2, 3),
                    btilde = c(-2.5, -2.5), mu_nd = c(0, 0.4),
                    rho = c(-0.05, -0.05), seed = 7)
fit <- mgmixirt(sim$resp, sim$group, model = "hybrid", quadpts = 9,
                starts = c(2, 6), maxit = 600, tol = 1e-4, seed = 1,
                verbose = FALSE)
fit
```

Person-level output — EAP abilities, posterior class probabilities, and
the paper's no-PD/PD classification rule:

```{r scores}
sc <- scores(fit)
head(sc, 3)
table(sc$pd_class, sc$group)
```

The cumulated decline-class probabilities per group (cf. List et al.,
2017, Fig. 2):

```{r classprob, fig.width = 6, fig.height = 4}
plot(fit, type = "classprob")
```

## Is there decline at all? Model comparison

The `"2pl"` baseline (no mixture) makes the question testable:

```{r ic}
fit0 <- mgmixirt(sim$resp, sim$group, model = "2pl", quadpts = 9,
                 starts = c(2, 6), maxit = 400, tol = 1e-4, seed = 1,
                 verbose = FALSE)
c(AIC_2pl = AIC(fit0), AIC_hybrid = AIC(fit))
c(BIC_2pl = BIC(fit0), BIC_hybrid = BIC(fit))
```

Note that BIC penalises the mixture heavily; in short tests with
moderate decline it can prefer the 2PL even when decline is present
(cf. Sen & Cohen, 2024). Reporting both criteria is good practice.

## Group comparisons with standard errors

`vcov()` computes the observed-information covariance matrix
numerically and `wald_test()` reports natural-scale group parameters
with standard errors and Wald tests of group equality (the analogue of
Table 2 in List et al., 2017):

```{r wald}
wt <- wald_test(fit)
print(wt, digits = 2)
```

## Cross-validation against Mplus

`mplus_syntax()` writes the full Mplus input for the fitted model —
the same specification that otherwise requires one model block per
group-by-class combination:

```{r mplus}
syntax <- mplus_syntax(fit, data_file = "mydata.dat")
cat(substr(syntax, 1, 420), "...\n")
```

The package's estimates have been validated against Mplus 8.3 on
simulated two-group 2PDM data (log-likelihoods agreeing to 0.03 units,
all identified parameters to < 0.03; see the package repository for the
validation artifacts).

## References

Bolt, D. M., Cohen, A. S., & Wollack, J. A. (2002). Item parameter
estimation under conditions of test speededness. *Journal of
Educational Measurement, 39*, 331–348.

Cao, J., & Stokes, S. L. (2008). Bayesian IRT guessing models for
partial guessing behaviors. *Psychometrika, 73*, 209–230.

Jin, K.-Y., & Wang, W.-C. (2014). Item response theory models for
performance decline during testing. *Journal of Educational
Measurement, 51*, 178–200.

List, M. K., Robitzsch, A., Lüdtke, O., Köller, O., & Nagy, G. (2017).
Performance decline in low-stakes educational assessments: Different
mixture modeling approaches. *Large-scale Assessments in Education, 5*,
15.

Yamamoto, K. (1995). *Estimating the effects of test length and test
time on parameter estimation using the HYBRID model* (ETS Research
Report RR-95-2). Educational Testing Service.
